# Data visualization in practice {#sec-dataviz-in-practice}

In this chapter, we will demonstrate how relatively simple **ggplot2** code can create insightful and aesthetically pleasing plots. As motivation we will create plots that help us better understand trends in global technology adoption. We will implement what we learned in @sec-ggplot2 and @sec-dataviz-principles and learn how to augment the code to perfect the plots. As we go through our case study, we will describe relevant general data visualization principles and learn concepts such as *faceting*, *time series plots*, *transformations*, and *ridge plots*.

## Case study 1: rethinking the digital divide

Sensationalist coverage of technology tends to pile dramatic stories into an over-dramatic worldview: "the digital divide is permanent", "it's the connected world versus everyone else", "some countries will never catch up". Effective data visualization is a strong antidote to this kind of narrative. In this section, we use data to attempt to answer the following two questions:

1.  Is it a fair characterization of today's world to say it is divided into a small set of mature digital markets and a much larger "digitally left behind" world in Africa, Asia, and Latin America?
2.  Has the gap in how much value each user generates (a market's "monetization intensity") across countries worsened during the last 40 years?

To answer these questions, we will be using the `global_tech_adoption` dataset provided in **behdslabs**. It reframes country-level indicators over time -- active user base, average app price, app uninstall rate, average daily screen time, and per-capita economic output. You can access the table like this:

```{r load-libraries, message=FALSE, echo=TRUE, cache=FALSE}
library(tidyverse)
library(behdslabs)
global_tech_adoption |> as_tibble()
```

We start by testing our intuition about differences in app uninstall rates across countries. For each of the five pairs of countries below, which country do you think had the higher app uninstall rate (uninstalls per 1,000 installs within 30 days) in 2015? Which pairs do you think are most similar?

1.  Sri Lanka or Turkey
2.  Poland or South Korea
3.  Malaysia or Russia
4.  Pakistan or Vietnam
5.  Thailand or South Africa

When answering these questions without data, the non-European countries are typically picked as having higher uninstall rates: Sri Lanka over Turkey, South Korea over Poland, and Malaysia over Russia. It is also common to assume that countries often lumped together as "emerging markets" -- Pakistan, Vietnam, Thailand, and South Africa -- have similarly high uninstall rates.

To answer these questions **with data**, we can use **dplyr**. For example, for the first comparison we see that:

```{r, message=FALSE}
global_tech_adoption |>
  filter(year == 2015 & country %in% c("Sri Lanka","Turkey")) |>
  select(country, app_uninstall_rate_per_1000)
```

Turkey has the higher uninstall rate.

We can use this code on all comparisons and find the following:

```{r, echo = FALSE}
comp_table <- tibble(comparison = rep(1:5, each = 2),
           country = c("Sri Lanka", "Turkey", "Poland", "South Korea", "Malaysia", "Russia", "Pakistan","Vietnam","Thailand","South Africa"))

tmp <- global_tech_adoption |>
  filter(year == 2015) |>
  select(country, app_uninstall_rate_per_1000) |>
  mutate(country = as.character(country)) ##to match characters to characters

tab <- inner_join(comp_table, tmp, by = "country") |> select(-comparison)

tmp <- cbind(slice(tab,seq(1,9,2)), slice(tab,seq(2,10,2)))
if (knitr::is_html_output()) {
  knitr::kable(tmp, "html") |>
    kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
} else{
  knitr::kable(tmp, "latex", booktabs = TRUE) |>
    kableExtra::kable_styling(font_size = 8)
}
```

We see that the European countries on this list have higher uninstall rates: Poland has a higher rate than South Korea, and Russia has a higher rate than Malaysia. We also see that Pakistan has a much higher rate than Vietnam, and South Africa has a much higher rate than Thailand. When this kind of quiz is given to educated audiences, the average score is typically worse than what they would have obtained by guessing at random. This suggests that we are not so much ignorant as misinformed. In this chapter we see how data visualization helps inform us.

## Scatterplots

The misconception described in the previous section stems from the preconceived notion that the world is divided into two groups: a set of mature digital markets (Western Europe and North America), characterized by heavy device use and cheap, freemium apps, versus a "digitally left behind" world (Africa, Asia, and Latin America) characterized by light device use and comparatively expensive apps. But do the data support this dichotomous view?

The necessary data to answer this question is also available in our `global_tech_adoption` table. Using our newly learned data visualization skills, we will be able to tackle this challenge.

In order to analyze this world view, our first plot is a scatterplot of average daily screen time versus average app price. We start by looking at data from about 50 years ago, when perhaps this view was first cemented in our minds.

```{r screentime-versus-price-1962, warning=FALSE}
filter(global_tech_adoption, year == 1962) |>
  ggplot(aes(avg_price_per_app_usd, avg_daily_screen_time_hours)) +
  geom_point()
```

Most points fall into two distinct categories:

1.  Daily screen time around 8 hours and average app prices of about \$3 or less.
2.  Daily screen time below about 7 hours and average app prices above \$5.

To confirm that indeed these countries are from the regions we expect, we can use color to represent continent.

```{r screentime-versus-price-1962-with-color}
filter(global_tech_adoption, year == 1962) |>
  ggplot( aes(avg_price_per_app_usd, avg_daily_screen_time_hours, color = continent)) +
  geom_point()
```

In 1962, "the mature markets versus everyone else" view was grounded in some reality. Is this still the case 50 years later?

## Faceting

We could easily plot the 2012 data in the same way we did for 1962. To make comparisons, however, side by side plots are preferable. In **ggplot2**, we can achieve this by *faceting* variables: we stratify the data by some variable and make the same plot for each strata.

To achieve faceting, we add a layer with the function `facet_grid`, which automatically separates the plots. This function lets you facet by up to two variables using columns to represent one variable and rows to represent the other. The function expects the row and column variables to be separated by a `~`. Here is an example of a scatterplot with `facet_grid` added as the last layer:

```{r screentime-versus-price-facet, warning=FALSE, out.width="100%", fig.height = 3}
filter(global_tech_adoption, year %in% c(1962, 2012)) |>
  ggplot(aes(avg_price_per_app_usd, avg_daily_screen_time_hours, col = continent)) +
  geom_point() +
  facet_grid(year~continent)
```

We see a plot for each continent/year pair. However, this is just an example and more than what we want, which is simply to compare 1962 and 2012. In this case, there is just one variable and we use `.` to let facet know that we are not using a second variable:

```{r screentime-versus-price-two-years, warning=FALSE, out.width="100%", fig.height=3}
filter(global_tech_adoption, year %in% c(1962, 2012)) |>
  ggplot(aes(avg_price_per_app_usd, avg_daily_screen_time_hours, col = continent)) +
  geom_point() +
  facet_grid(. ~ year)
```

This plot clearly shows that the majority of countries have moved from the *lagging* cluster to the *mature market* one. In 2012, the "mature markets versus everyone else" view no longer makes sense. This is particularly clear when comparing Europe to Asia, the latter of which includes several countries that have caught up dramatically.

### `facet_wrap`

To explore how this transformation happened through the years, we can make the plot for several years. For example, we can add 1970, 1980, 1990, and 2000. If we do this, we will not want all the plots on the same row, the default behavior of `facet_grid`, since they will become too thin to show the data. Instead, we will want to use multiple rows and columns. The function `facet_wrap` permits us to do this by automatically wrapping the series of plots so that each display has viewable dimensions:

```{r screentime-versus-price-five-years, out.width="100%"}
years <- c(1962, 1980, 1990, 2000, 2012)
continents <- c("Europe", "Asia")
global_tech_adoption |>
  filter(year %in% years & continent %in% continents) |>
  ggplot( aes(avg_price_per_app_usd, avg_daily_screen_time_hours, col = continent)) +
  geom_point() +
  facet_wrap(~year)
```

This plot clearly shows how most Asian markets have matured at a much faster rate than European ones.

### Fixed scales for better comparisons

The default choice of the range of the axes is important. When not using `facet`, this range is determined by the data shown in the plot. When using `facet`, this range is determined by the data shown in all plots and therefore kept fixed across plots. This makes comparisons across plots much easier. For example, in the above plot, we can see that screen time has increased and app prices have decreased across most countries. We see this because the cloud of points moves. This is not the case if we adjust the scales:

```{r facet-without-fixed-scales, warning=FALSE}
filter(global_tech_adoption, year %in% c(1962, 2012)) |>
  ggplot(aes(avg_price_per_app_usd, avg_daily_screen_time_hours, col = continent)) +
  geom_point() +
  facet_wrap(. ~ year, scales = "free")
```

In the plot above, we have to pay special attention to the range to notice that the plot on the right has larger screen time.

## Time series plots

The visualizations above effectively illustrate that data no longer supports the "mature markets versus everyone else" view. Once we see these plots, new questions emerge. For example, which countries are maturing more and which ones less? Was the change constant during the last 50 years or was it more accelerated during certain periods? For a closer look that may help answer these questions, we introduce *time series plots*.

Time series plots have time in the x-axis and an outcome or measurement of interest on the y-axis. For example, here is a trend plot of the United States' average app price:

```{r price-time-series-plot-points, warning=FALSE}
global_tech_adoption |>
  filter(country == "United States") |>
  ggplot(aes(year, avg_price_per_app_usd)) +
  geom_point()
```

We see that the trend is not linear at all. Instead there is a sharp drop during the 1960s and 1970s to below \$2. Then the trend comes back to about \$2 and stabilizes during the 1990s.

When the points are regularly and densely spaced, as they are here, we create curves by joining the points with lines, to convey that these data are from a single series, here a country. To do this, we use the `geom_line` function instead of `geom_point`.

```{r price-time-series-plot-curve, warning=FALSE}
global_tech_adoption |>
  filter(country == "United States") |>
  ggplot(aes(year, avg_price_per_app_usd)) +
  geom_line()
```

This is particularly helpful when we look at two countries. We can subset the data to include two countries, one from Europe and one from Asia, then adapt the code above:

```{r wrong-time-series-plot, warning=FALSE, message=FALSE}
countries <- c("South Korea", "Germany")

global_tech_adoption |> filter(country %in% countries) |>
  ggplot(aes(year, avg_price_per_app_usd)) +
  geom_line()
```

Unfortunately, this is **not** the plot that we want. Rather than a line for each country, the points for both countries are joined. This is actually expected since we have not told `ggplot` anything about wanting two separate lines. To let `ggplot` know that there are two curves that need to be made separately, we assign each point to a `group`, one for each country:

```{r time-series-two-curves, warning=FALSE, message=FALSE}
countries <- c("South Korea","Germany")

global_tech_adoption |> filter(country %in% countries & !is.na(avg_price_per_app_usd)) |>
  ggplot(aes(year, avg_price_per_app_usd, group = country)) +
  geom_line()
```

But which line goes with which country? We can assign colors to make this distinction. A useful side-effect of using the `color` argument to assign different colors to the different countries is that the data is automatically grouped:

```{r price-time-series-plot}
countries <- c("South Korea","Germany")
global_tech_adoption |> filter(country %in% countries & !is.na(avg_price_per_app_usd)) |>
  ggplot(aes(year, avg_price_per_app_usd, col = country)) +
  geom_line()
```

The plot clearly shows how South Korea's average app price fell drastically during the 1960s and 1970s, and by 1990 had converged to a level similar to Germany's.

For trend plots we recommend labeling the lines rather than using legends since the viewer can quickly see which line is which country. This suggestion actually applies to most plots: labeling is usually preferred over legends.

We demonstrate how we can do this using the `geomtextpath` package. We define a data table with the label locations and then use a second mapping just for these labels:

```{r labels-better-than-legends}
library(geomtextpath)
global_tech_adoption |>
  filter(country %in% countries) |>
  ggplot(aes(year, avg_daily_screen_time_hours, col = country, label = country)) +
  geom_textpath() +
  theme(legend.position = "none")
```

The plot clearly shows how a rise in screen time followed the drop in app prices. In 1960, Germans averaged about 2 more hours of daily screen time than South Koreans, although by 2010 the gap is completely closed. It exemplifies the catch-up that many emerging markets have achieved in the last 40 years.

## Data transformations

We now shift our attention to the second question related to the commonly held notion that the gap in monetization intensity across the world has become worse during the last decades. When general audiences are asked whether low-revenue markets have become even less lucrative and high-revenue markets even more so, the majority answers yes. By using stratification, histograms, smooth densities, and boxplots, we will be able to understand if this is in fact the case. First we learn how transformations can sometimes help provide more informative summaries and plots.

The `global_tech_adoption` data table includes a column with the countries' gross domestic product (GDP), a rough summary of a country's economic output. Here we divide GDP by the active user base and by 365 to obtain a more interpretable measure: *revenue per user per day*, a proxy for how much economic value each active user in a market represents daily. A market where this figure is below \$2 a day is one where digital services are, on average, barely monetized. We add this variable to the data table:

```{r}
global_tech_adoption <- global_tech_adoption |>
  mutate(revenue_per_user_day = gdp/active_user_base/365)
```

The GDP values are adjusted for inflation and represent current US dollars, so these values are meant to be comparable across the years. Of course, these are country averages and within each country there is much variability. All the graphs and insights described below relate to country averages and not to individuals.

### Log transformation

Here is a histogram of revenue per user per day from 1970:

```{r dollars-per-day-distribution}
past_year <- 1970
global_tech_adoption |>
  filter(year == past_year & !is.na(gdp)) |>
  ggplot(aes(revenue_per_user_day)) +
  geom_histogram(binwidth = 1, color = "black")
```

We use the `color = "black"` argument to draw a boundary and clearly distinguish the bins.

In this plot, we see that for the majority of countries, averages are below \$10 a day. However, the majority of the x-axis is dedicated to the `r filter(global_tech_adoption, year == past_year & !is.na(gdp)) |> summarise(x = sum(revenue_per_user_day>10)) |> pull(x)` countries with averages above \$10. So the plot is not very informative about countries with values below \$10 a day.

It might be more informative to quickly be able to see how many countries have a revenue per user per day of about \$1 (very low monetization), \$2, \$4, \$8, \$16, \$32, \$64 (very high monetization) per day. These changes are multiplicative, and log transformations convert multiplicative changes into additive ones: when using base 2, a doubling of a value turns into an increase by 1.

Here is the distribution if we apply a log base 2 transform:

```{r dollars-per-day-distribution-log}
global_tech_adoption |>
  filter(year == past_year & !is.na(gdp)) |>
  ggplot(aes(log2(revenue_per_user_day))) +
  geom_histogram(binwidth = 1, color = "black")
```

In a way, this provides a *close-up* of the mid to lower income countries.

### Which base?

In the case above, we used base 2 in the log transformations. Other common choices are base $\mathrm{e}$ (the natural log) and base 10.

In general, we do not recommend using the natural log for data exploration and visualization. This is because while $2^2, 2^3, 2^4, \dots$ or $10^2, 10^3, \dots$ are easy to mentally compute, but the same is not true for $\mathrm{e}^2, \mathrm{e}^3, \dots$. The natural log scale is not intuitive or easy to interpret.

In the revenue per user example, we used base 2 instead of base 10 because the resulting range is easier to interpret. The range of the untransformed values is `r with(filter(global_tech_adoption, year==past_year), range(revenue_per_user_day, na.rm=TRUE))`.

In base 10, this turns into a range that includes very few integers: just 0 and 1. With base 2, our range includes -1, 0, 1, 2, 3, 4, and 5. It is easier to compute $2^x$ and $10^x$ when $x$ is an integer and between -10 and 10, so we prefer to have smaller integers in the scale. Another consequence of a limited range is that choosing the binwidth is more challenging. With log base 2, we know that a binwidth of 1 will translate to a bin with range $x$ to $2x$.

For an example in which base 10 makes more sense, consider active user bases. A log base 10 is preferable since the range for these is:

```{r}
filter(global_tech_adoption, year == past_year) |>
  summarize(min = min(active_user_base), max = max(active_user_base))
```

Here is the histogram of the transformed values:

```{r population-histogram-log10}
global_tech_adoption |>
  filter(year == past_year) |>
  ggplot(aes(log10(active_user_base))) +
  geom_histogram(binwidth = 0.5, color = "black")
```

In the above, we quickly see that country active user bases range between about ten thousand and one billion.

### Transform the values or the scale?

There are two ways we can use log transformations in plots. We can log the values before plotting them or use log scales in the axes. The plot will look the same,  except for the numbers in the axes. Both approaches are useful and have different strengths. If we log the data, we can more easily interpret intermediate values in the scale. For example, if we see:

`----1----x----2--------3----`

for log transformed data, we know that the value of $x$ is about 1.5. If the scales are logged:

`----10---x---100------1000---`

then, to determine `x`, we need to compute $10^{1.5}$, which is not easy to do in our heads. However, the advantage of showing logged scales is that the original values are displayed in the plot, which are easier to interpret. For example, we would see "32 dollars a day" instead of "5 log base 2 dollars a day".

As we learned earlier, if we want to scale the axis with logs, we can use the `scale_x_continuous` function. Instead of logging the values first, we apply this layer:

```{r dollars-per-day-log-scale}
global_tech_adoption |>
  filter(year == past_year & !is.na(gdp)) |>
  ggplot(aes(revenue_per_user_day)) +
  geom_histogram(binwidth = 1, color = "black") +
  scale_x_continuous(trans = "log2")
```

Note that the log base 10 transformation has its own function: `scale_x_log10()`, but currently base 2 does not, although we could easily define our own.

There are other transformations available through the `trans` argument. As we learn later on, the square root (`sqrt`) transformation is useful when considering counts. The logistic transformation (`logit`) is useful when plotting proportions between 0 and 1. The `reverse` transformation is useful when we want smaller values to be on the right or on top.

## Multimodal distributions

In the histogram above we see two *bumps*: one at about 4 and another at about 32. In statistics these bumps are sometimes referred to as *modes*. The mode of a distribution is the value with the highest frequency. The mode of the normal distribution is the average. When a distribution, like the one above, doesn't monotonically decrease from the mode, we call the locations where it goes up and down again *local modes* and say that the distribution has *multiple modes*.

The histogram above suggests that the 1970 revenue-per-user distribution has two modes: one at about \$2 per user per day (1 in the log 2 scale) and another at about \$32 per user per day (5 in the log 2 scale). This *bimodality* is consistent with a dichotomous world made up of markets generating less than \$8 per user per day (3 in the log 2 scale) and markets above that.

## Comparing distributions

A histogram showed us that the 1970 revenue-per-user distribution shows a dichotomy. However, the histogram does not show us if the two groups of countries are *mature markets* versus the rest.

Let's start by quickly examining the data by region. We reorder the regions by the median value and use a log scale.

```{r dollars-per-day-points}
global_tech_adoption |>
  filter(year == past_year & !is.na(gdp)) |>
  mutate(region = reorder(region, revenue_per_user_day, FUN = median)) |>
  ggplot(aes(revenue_per_user_day, region)) +
  geom_point() +
  scale_x_continuous(trans = "log2")
```

We can already see that there is indeed a "mature markets versus the rest" dichotomy: we see two clear groups, with the high-revenue group composed of North America, Northern and Western Europe, and New Zealand and Australia. We define groups based on this observation:

```{r}
global_tech_adoption <- global_tech_adoption |>
  mutate(group = case_when(
    region %in% c("Western Europe", "Northern Europe","Southern Europe", 
                    "Northern America", 
                  "Australia and New Zealand") ~ "West",
    region %in% c("Eastern Asia", "South-Eastern Asia") ~ "East Asia",
    region %in% c("Caribbean", "Central America", 
                  "South America") ~ "Latin America",
    continent == "Africa" & 
      region != "Northern Africa" ~ "Sub-Saharan",
    TRUE ~ "Others"))
```

We turn this `group` variable into a factor to control the order of the levels:

```{r}
global_tech_adoption <- global_tech_adoption |>
  mutate(group = factor(group, levels = c("Others", "Latin America", 
                                          "East Asia", "Sub-Saharan",
                                          "West")))
```

In the next section we demonstrate how to visualize and compare distributions across groups.

### Boxplots

The exploratory data analysis above has revealed two characteristics about the revenue-per-user distribution in 1970. Using a histogram, we found a bimodal distribution with the modes relating to low- and high-monetization markets. We now want to compare the distribution across these five groups to confirm the "mature markets versus the rest" dichotomy. The number of points in each category is large enough that a summary plot may be useful. We could generate five histograms or five density plots, but it may be more practical to have all the visual summaries in one plot. We therefore start by stacking boxplots next to each other. Note that we add the layer `theme(axis.text.x = element_text(angle = 90, hjust = 1))` to turn the group labels vertical, since they do not fit if we show them horizontally, and we remove the axis label to make space.

```{r dollars-per-day-boxplot}
p <- global_tech_adoption |>
  filter(year == past_year & !is.na(gdp)) |>
  ggplot(aes(group, revenue_per_user_day)) +
  geom_boxplot() +
  scale_y_continuous(trans = "log2") +
  xlab("") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
p
```

Boxplots have the limitation that by summarizing the data into five numbers, we might miss important characteristics of the data. One way to avoid this is by showing the data.

```{r dollars-per-day-boxplot-with-data}
p + geom_point(alpha = 0.5)
```

### Ridge plots

Showing each individual point does not always reveal important characteristics of the distribution. Although not the case here, when the number of data points is so large that there is over-plotting, showing the data can be counterproductive. Boxplots help with this by providing a five-number summary, but this has limitations too. For example, boxplots will not permit us to discover bimodal distributions. To see this, note that the two plots below are summarizing the same dataset:

```{r boxplot-dont-show-bimodal, warning=FALSE, out.width="100%", message = FALSE, echo=FALSE, fig.height=3}
set.seed(1987)
z <- sample(c(0,1), 1000, replace = TRUE, prob = c(0.25, 0.75))
x <- rnorm(100)*z + rnorm(100, 5)*(1 - z)
p1 <- data.frame(x=x) |> 
  ggplot(aes(x)) + 
  geom_density(fill = 1, show.legend=FALSE, alpha = 0.2) +
  scale_x_continuous(limits=c(-4,8.5))
p2 <- data.frame(g= "", x=x) |> ggplot(aes(g, x)) + geom_boxplot() + xlab("") +
  theme(axis.ticks = element_blank())
  
gridExtra::grid.arrange(p1, p2, nrow = 1)
```

In cases in which we are concerned that the boxplot summary is too simplistic, we can show stacked smooth densities or histograms. We refer to these as *ridge plots*. Because we are used to visualizing densities with values in the x-axis, we stack them vertically. Also, because more space is needed in this approach, it is convenient to overlay them. The package **ggridges** provides a convenient function for doing this. Here is the income data shown above with boxplots but with a *ridge plot*.

```{r ridge-plot, message=FALSE}
library(ggridges)
p <- global_tech_adoption |>
  filter(year == past_year & !is.na(revenue_per_user_day)) |>
  ggplot(aes(revenue_per_user_day, group)) + 
  scale_x_continuous(trans = "log2") 
p  + geom_density_ridges() 
```

Note that we have to invert the `x` and `y` used for the boxplot. A useful `geom_density_ridges` parameter is `scale`, which lets you determine the amount of overlap, with `scale = 1` meaning no overlap and larger values resulting in more overlap.

If the number of data points is small enough, we can add them to the ridge plot using the following code:

```{r ridge-plot-with-points, message=FALSE}
p + geom_density_ridges(jittered_points = TRUE)
```

By default, the height of the points is jittered and should not be interpreted in any way. To show data points, but without using jitter we can use the following code to add what is referred to as a *rug representation* of the data.

```{r ridge-plot-with-rug, message=FALSE}
p + geom_density_ridges(jittered_points = TRUE, 
                        position = position_points_jitter(height = 0),
                        point_shape = '|', point_size = 3, 
                        point_alpha = 1, alpha = 0.7)
```

### Example: 1970 versus 2010 revenue-per-user distributions

Data exploration clearly shows that in 1970 there was a "mature markets versus the rest" dichotomy. But does this dichotomy persist? Let's use `facet_grid` and see how the distributions have changed. To start, we will focus on two groups: the mature markets and the rest. We make four histograms. We make this plot only for countries with data in both 1970 and 2010. Note that several countries were founded after 1970; for example, the Soviet Union divided into several countries during the 1990s. We also note that that data was available for more countries in 2010.

We therefore make the plot only for countries with data in both years:
```{r}
past_year <- 1970
present_year <- 2010
years <- c(past_year, present_year)
country_list <- global_tech_adoption |>
  filter(year %in% c(present_year, past_year)) |>
  group_by(country) |>
  summarize(n = sum(!is.na(revenue_per_user_day)), .groups = "drop") |>
  filter(n == 2) |>
  pull(country)
```

These `r length(country_list)` countries account for `r round(global_tech_adoption |>filter(year==present_year) |> summarize(perc=sum(active_user_base[country%in%country_list], na.rm=TRUE)/sum(active_user_base, na.rm=TRUE)) |> pull(perc)*100 )`% of the world's active user base, so this subset should be representative. We can compare the distributions using this code:


```{r income-histogram-west-v-devel}
global_tech_adoption |>
  filter(year %in% years & country %in% country_list) |>
  mutate(west = ifelse(group == "West", "Mature", "Emerging")) |>
  ggplot(aes(revenue_per_user_day)) +
  geom_histogram(binwidth = 1, color = "black") +
  scale_x_continuous(trans = "log2") +
  facet_grid(year ~ west)
```

We now see that the high-revenue markets have grown a bit more, but percentage-wise, the lower-revenue markets appear to have grown more. In particular, we see that the proportion of *emerging* markets generating more than \$16 per user per day increased substantially.

To see which specific regions changed the most, we can remake the boxplots we made above, but now adding the year 2010 and then using facet to compare the two years.

```{r  income-histogram-by-region, out.width="100%"}
global_tech_adoption |>
  filter(year %in% years & country %in% country_list) |>
  ggplot(aes(group, revenue_per_user_day)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_y_continuous(trans = "log2") +
  xlab("") +
  facet_grid(. ~ year)
```

Here, we pause to introduce another powerful **ggplot2** feature. Because we want to compare each region before and after, it would be convenient to have the `r past_year` boxplot next to the `r present_year` boxplot for each region. In general, comparisons are easier when data are plotted next to each other.

So instead of faceting, we keep the data from each year together and ask to color (or fill) them depending on the year. Note that groups are automatically separated by year and each pair of boxplots drawn next to each other. Because the year column is a number, we have to convert it into a factor since **ggplot2** automatically assigns a color to each category of a factor.

```{r income-histogram-west-v-devel-by-year}
global_tech_adoption |>
  filter(year %in% years & country %in% country_list) |>
  mutate(year = factor(year)) |>
  ggplot(aes(group, revenue_per_user_day, fill = year)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_y_continuous(trans = "log2") +
  xlab("") 
```


The previous data exploration suggested that the revenue-per-user gap between high- and low-monetization markets has narrowed considerably during the last 40 years. We used a series of histograms and boxplots to see this. We suggest a succinct way to convey this message with just one plot.

Let's start by noting that density plots for the revenue-per-user distribution in `r past_year` and `r present_year` deliver the message that the gap is closing:

```{r income-smooth-density-by-year, out.width="100%", fig.height=3}
global_tech_adoption |>
  filter(year %in% years & country %in% country_list) |>
  ggplot(aes(revenue_per_user_day)) +
  geom_density(fill = "grey") + 
  scale_x_continuous(trans = "log2") + 
  facet_grid(. ~ year)
```

In the `r past_year` plot, we see two clear modes: low- and high-monetization markets. In `r present_year`, it appears that some of the low-monetization markets have shifted towards the right, closing the gap.

The next message we need to convey is that the reason for this change in distribution is that several low-monetization markets moved up, rather than some high-monetization markets moving down. To do this, we can assign a color to the groups we identified during data exploration.

However, because when we overlay two densities, the default is to have the area under the distribution curve add up to 1 for each group, regardless of the size of each group, we first need to learn how to make these smooth densities in a way that preserves information on the number of countries in each group. To do this, we will need to learn to access computed variables with the `geom_density` function.

### Accessing computed variables

To have the areas of these densities be proportional to the size of the groups, we can simply multiply the y-axis values by the size of the group. From the `geom_density` help file, we see that the functions compute a variable called `count` that does exactly this. We want this variable to be on the y-axis rather than the density.

In **ggplot2**, we access these variables using the function `after_stat`. We will therefore use the following mapping:

```{r, eval=FALSE}
aes(x = revenue_per_user_day, y = after_stat(count))
```

We can now create the desired plot by simply changing the mapping in the previous code chunk. We will also expand the limits of the x-axis.

```{r income-smooth-density-counts}
p <- global_tech_adoption |>
  filter(year %in% years & country %in% country_list) |>
  mutate(group = ifelse(group == "West", "Mature", "Emerging")) |>
  ggplot(aes(revenue_per_user_day, y = after_stat(count), fill = group)) +
  scale_x_continuous(trans = "log2", limits = c(0.125, 300))
p + geom_density(alpha = 0.2) + facet_grid(year ~ .)
```

If we want the densities to be smoother, we use the `bw` argument so that the same bandwidth is used in each density. We selected 0.75 after trying out several values.

```{r income-smooth-density-counts-by-year}
p + geom_density(alpha = 0.2, bw = 0.75) + facet_grid(year ~ .)
```

This plot now shows what is happening very clearly. The emerging-market distribution is changing. A third mode appears consisting of the markets that most narrowed the gap.

To visualize if any of the groups defined above are driving this we can quickly make a ridge plot:

```{r ridge-plot-income-five-regions, warning=FALSE, message=FALSE}
global_tech_adoption |>
  filter(year %in% years & !is.na(revenue_per_user_day)) |>
  ggplot(aes(revenue_per_user_day, group)) + 
  scale_x_continuous(trans = "log2") + 
  geom_density_ridges(bandwidth = 1.5) +
  facet_grid(. ~ year)
```

Another way to achieve this is by stacking the densities on top of each other:

```{r income-smooth-density-counts-by-region-and-year}
global_tech_adoption |>
    filter(year %in% years & country %in% country_list) |>
  group_by(year) |>
  mutate(weight = active_user_base/sum(active_user_base)*2) |>
  ungroup() |>
  ggplot(aes(revenue_per_user_day, fill = group)) +
  scale_x_continuous(trans = "log2", limits = c(0.125, 300)) + 
  geom_density(alpha = 0.2, bw = 0.75, position = "stack") + 
  facet_grid(year ~ .) 
```

Here we can clearly see how the distributions for East Asia, Latin America, and others shift markedly to the right while Sub-Saharan Africa remains stagnant.

Notice that we order the levels of the group so that the West's density is plotted first, then Sub-Saharan Africa. Having the two extremes plotted first allows us to see the remaining bimodality better.

### Weighted densities

As a final point, we note that these distributions weigh every country the same. So if most users are in markets that are improving, but concentrated in a very large market, such as China, we might not appreciate this. We can actually weight the smooth densities using the `weight` mapping argument. The plot then looks like this:

```{r income-smooth-density-counts-by-region-year-weighted, warning=FALSE, echo=FALSE}
global_tech_adoption |>
  filter(year %in% years & country %in% country_list) |>
  group_by(year) |>
  mutate(weight = active_user_base/sum(active_user_base)*2) |>
  ungroup() |>
  ggplot(aes(revenue_per_user_day, fill = group, weight = weight)) +
  scale_x_continuous(trans = "log2", limits = c(0.125, 300)) + 
  geom_density(alpha = 0.2, bw = 0.75, position = "stack") + facet_grid(year ~ .) 
```

This particular figure shows very clearly how the revenue-per-user gap is closing, with the lowest-revenue markets remaining in Sub-Saharan Africa.

## Case study 2: the ecological fallacy

Throughout this section, we have been comparing regions of the world. We have seen that, on average, some regions do better than others. In this section, we focus on describing the importance of variability within the groups when examining the relationship between a country's app uninstall rate and its revenue per user.

We define a few more regions and compare the averages across regions:

```{r ecological-fallacy-averages, echo=FALSE}
library(ggrepel)
global_tech_adoption <- global_tech_adoption |>
  mutate(group = case_when(
    region %in% c("Western Europe", "Northern Europe",
                  "Southern Europe", "Northern America",
                  "Australia and New Zealand") ~ "West",
    region %in% "Northern Africa" ~ "Northern Africa",
    region %in% c("Eastern Asia", "South-Eastern Asia") ~ "East Asia",
    region == "Southern Asia"~ "Southern Asia",
    region %in% c("Central America", "South America",
                  "Caribbean") ~ "Latin America",
    continent == "Africa" &
      region != "Northern Africa" ~ "Sub-Saharan",
    region %in% c("Melanesia", "Micronesia",
                  "Polynesia") ~ "Pacific Islands"))
surv_income <- global_tech_adoption |>
  filter(year %in% present_year & !is.na(gdp) &
           !is.na(app_uninstall_rate_per_1000) & !is.na(group)) |>
  group_by(group) |>
  summarize(revenue = sum(gdp)/sum(active_user_base)/365,
            retention_rate =
              1 - sum(app_uninstall_rate_per_1000/1000*active_user_base)/sum(active_user_base))

#surv_income |> arrange(revenue) |> print(n=nrow(surv_income))

surv_income |> ggplot(aes(revenue, retention_rate, label = group, color = group)) +
  scale_x_continuous(trans = "log2", limits = c(0.25, 150)) +
  scale_y_continuous(trans = "logit", limits = c(0.875, .9981),
                     breaks = c(.85,.90,.95,.99,.995,.998)) +
  geom_label_repel(size = 3, show.legend = FALSE)
```

The relationship between these two variables is almost perfectly linear with the chosen axis transformations, and the graph shows a dramatic difference. While in the West only about 0.5% of installs are uninstalled within the tracked window, in Sub-Saharan Africa the rate is higher than 6%!

Note that the plot uses a new transformation, the logistic transformation.

### Logistic transformation {#sec-logit}

The logistic or logit transformation for a proportion or rate $p$ is defined as:

$$f(p) = \log \left( \frac{p}{1-p} \right)$$

When $p$ is a proportion or probability, the quantity that is being logged, $p/(1-p)$, is called the *odds*. In this case $p$ is the proportion of installs that were retained. The odds tell us how many more installs are expected to be retained than uninstalled. The log transformation makes this symmetric. If the rates are the same, then the log odds is 0. Fold increases or decreases turn into positive and negative increments, respectively.

This scale is useful when we want to highlight differences near 0 or 1. For retention proportions this is important because the difference between 0.90 and 0.99 matters far more than the raw gap suggests, and 0.999 is better still. We want our scale to highlight these differences and the logit does this. Note that 99.9/0.1 is about 10 times bigger than 99/1 which is about 10 times larger than 90/10. By using the log, these fold changes turn into constant increases.

### Show the data

Now, back to our plot. Based on the plot above, do we conclude that a market with low revenue per user is destined to have a low retention rate? Do we conclude that retention rates in Sub-Saharan Africa are all lower than in Southern Asia, which in turn are lower than in the Pacific Islands, and so on?

Jumping to this conclusion based on a plot showing averages is referred to as the *ecological fallacy*. The almost perfect relationship between retention rates and revenue per user is only observed for the averages at the region level. Once we show all the data, we see a somewhat more complicated story:

```{r ecological-fallacy-all-data, echo=FALSE}
library(ggrepel)
highlight <- c("Sierra Leone", "Mauritius",  "Sudan", "Botswana", "Tunisia",
               "Cambodia","Singapore","Chile", "Haiti", "Bolivia",
               "United States","Sweden", "Angola", "Serbia")

global_tech_adoption |>filter(year %in% present_year & !is.na(gdp) & !is.na(app_uninstall_rate_per_1000) & !is.na(group) ) |>
  mutate(country_name = ifelse(country %in% highlight, as.character(country), "")) |>
  ggplot(aes(revenue_per_user_day, 1 - app_uninstall_rate_per_1000/1000, col = group, label = country_name)) +
  scale_x_continuous(trans = "log2", limits = c(0.25, 150)) +
  scale_y_continuous(trans = "logit",limits = c(0.875, .9981),
                     breaks = c(.85,.90,.95,.99,.995,.998)) +
  geom_point(alpha = 0.5, size = 3) +
  geom_text_repel(size = 4, show.legend = FALSE)
```

Specifically, we see that there is a large amount of variability. We see that countries from the same regions can be quite different and that countries with the same revenue per user can have different retention rates. For example, while on average Sub-Saharan Africa had the worst retention and revenue outcomes, there is wide variability within that group. Mauritius and Botswana retain users better than Angola and Sierra Leone, with Mauritius comparable to Western countries.

## Case study 3: misinformation and app adoption {#sec-vaccines}


Guided-meditation and other digital wellbeing apps have helped millions of people manage stress, anxiety, and sleep. Before smartphones made these tools cheap and ubiquitous, structured relaxation training was largely confined to clinics and specialist courses. Today, however, digital wellbeing tools have become controversial in some circles, despite the evidence for their benefits.

Around 1963, as phone makers began bundling a free guided-meditation feature directly into their operating systems, downloads of the standalone apps started to fall. The decline was then amplified by a widely shared claim, later traced to a coordinated campaign, that guided-meditation apps induced dependency, worsened anxiety, and secretly harvested biometric data to sell to insurers.
Despite much evidence contradicting these claims, sensationalist posts and fear-mongering from a handful of accounts led parts of the public to believe that the apps were harmful. As a result, many people stopped using them altogether, and a wave of one-star review-bombing pushed the standalone apps further down the app-store rankings. The original claims have since been debunked and the campaign's coordinating accounts exposed and removed, yet the misconceptions persist, in part due to self-proclaimed activists who continue to disseminate misinformation about digital wellbeing tools.

Effective communication of data is a strong antidote to misinformation and fear-mongering. In the introduction to this part of the book we showed an example, in the style of a Wall Street Journal infographic^[<http://graphics.wsj.com/infectious-diseases-and-vaccines/>], showing state-by-state download data for a fitness-app category over decades. Here we reconstruct that example.


### App-download data

The data used for these plots are weekly-tracked download counts for seven fitness/wellness app categories from 1928 to 2011, from all fifty states. We include the yearly totals in the __behdslabs__ package:

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(RColorBrewer)
library(behdslabs)
names(fitness_app_downloads_by_state)
```

We create a temporary object `dat` that stores only the Guided Meditation data, includes a per 10,000 rate, orders states by average download rate and removes Alaska and Hawaii since they only became states in the late 1950s. Note that there is a `weeks_tracked` column that tells us for how many weeks of the year download data was tracked. We have to adjust for that value when computing the rate.

```{r}
the_category <- "Guided Meditation"
dat <- fitness_app_downloads_by_state |>
  filter(!state %in% c("Hawaii","Alaska") & app_category == the_category) |>
  mutate(rate = downloads / population * 10000 * 52 / weeks_tracked) |>
  mutate(state = reorder(state, ifelse(year <= 1963, rate, NA),
                         median, na.rm = TRUE))
```


### Trend plots

We can now easily plot download rates per year. Here are the Guided Meditation data from California:

```{r california-meditation-time-series}
dat |> filter(state == "California" & !is.na(rate)) |>
  ggplot(aes(year, rate)) +
  geom_line() +
  ylab("Downloads per 10,000")  +
  geom_vline(xintercept = 1963, col = "blue")
```

We add a vertical line at 1963, marking both the bundling of the free OS feature and the point at which the misinformation campaign against guided-meditation apps went viral.


## Heatmaps

Now can we show data for all states in one plot? We have three variables to show: year, state, and rate. In the infographic, the x-axis is used for year, the y-axis for state, and color hue to represent rates. However, a color scale that goes from yellow to blue to green to orange to red can be improved.

In our example, we want to use a sequential palette since there is no meaningful center, just low and high rates.

We use the geometry `geom_tile` to tile the region with colors representing download rates. We use a square root transformation to avoid having the really high counts dominate the plot. Notice that missing values are shown in grey. Note that once a category's standalone apps had mostly faded, some states stopped tracking downloads all together. This is why we see so much grey after 1980.

```{r downloads-heatmap, out.width="100%", fig.height=5}
dat |> ggplot(aes(year, state, fill = rate)) +
  geom_tile(color = "grey50") +
  scale_x_continuous(expand = c(0,0)) +
  scale_fill_gradientn(colors = brewer.pal(9, "Reds"), trans = "sqrt") +
  geom_vline(xintercept = 1963, col = "blue") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        legend.position = "bottom",
        text = element_text(size = 8)) +
  labs(title = the_category, x = "", y = "")
```

This plot makes a very striking case for how quickly the campaign and the platform shift reshaped the market. However, one limitation of this plot is that it uses color to represent quantity, which we earlier explained makes it harder to know exactly how high values are going. Position and lengths are better cues. If we are willing to lose state information, we can make a version of the plot that shows the values with position. We can also show the average for the US, which we compute like this:

```{r}
avg <- fitness_app_downloads_by_state |>
  filter(app_category == the_category) |> group_by(year) |>
  summarize(us_rate = sum(downloads, na.rm = TRUE) /
              sum(population, na.rm = TRUE) * 10000)
```

Now to make the plot we simply use the `geom_line` geometry:
```{r downloads-time-series}
dat |>
  filter(!is.na(rate)) |>
    ggplot() +
  geom_line(aes(year, rate, group = state),  color = "grey50",
            show.legend = FALSE, alpha = 0.2, linewidth = 1) +
  geom_line(mapping = aes(year, us_rate),  data = avg, linewidth = 1) +
  scale_y_continuous(trans = "sqrt", breaks = c(5, 25, 125, 300)) +
  ggtitle("Downloads per 10,000 by state") +
  xlab("") + ylab("") +
  geom_text(data = data.frame(x = 1955, y = 50),
            mapping = aes(x, y, label = "US average"),
            color = "black") +
  geom_vline(xintercept = 1963, col = "blue")
```

In theory, we could use color to represent the categorical value state, but it is hard to pick 50 distinct colors.

## Exercises


1. Reproduce the image plot we previously made but for the Step Tracking category. For this plot, do not include years in which downloads were not tracked in 10 or more weeks.

2. Now reproduce the time series plot we previously made, but this time following the instructions of the previous question for Step Tracking.

3. For the state of California, make a time series plot showing rates for all app categories. Include only years with 10 or more weeks tracked. Use a different color for each category.

4. Now do the same for the rates for the US. Hint: compute the US rate as the total divided by total population using `summarize`.



