# Text analysis

With the exception of labels used to represent categorical data, we have focused on numerical data. But in many applications, data starts as text. Well-known examples are spam filtering, cyber-crime prevention, counter-terrorism, and sentiment analysis. In all these cases, the raw data is composed of free-form text. Our task is to extract insights from these data. In this section, we learn how to generate useful numerical summaries from text data to which we can apply some of the powerful data visualization and analysis techniques we have learned.

## Case study: human versus AI answers

Since late 2022, large language models (LLMs) such as ChatGPT can produce fluent answers to almost any question. This raises a practical question for anyone working with user-generated text -- product reviews, survey responses, forum posts, homework: **can we tell, from the text alone, whether a human or an LLM wrote it?** This matters for detecting fake reviews, for flagging AI-assisted exam answers, and, more generally, for understanding how machine-generated text differs from the human kind.

To study this we use the **Human ChatGPT Comparison Corpus (HC3)**[^text-analysis-1], introduced in a 2023 paper by Guo and co-authors[^text-analysis-2]. For each of thousands of questions the corpus provides one or more answers written by people (often domain experts, or highly rated forum answers) *and* one or more answers generated by ChatGPT for the same question. Because the question is held fixed, any systematic difference we find between the two sets of answers is a difference in *authorship*, not in topic. This is the same idea behind a classic text-analysis case study[^text-analysis-3] that asked whether two different people were posting from a single social-media account by comparing the language coming from each device.

[^text-analysis-1]: <https://huggingface.co/datasets/Hello-SimpleAI/HC3>. Released under a CC BY-SA 4.0 license.

[^text-analysis-2]: Guo et al. (2023), "How Close is ChatGPT to Human Experts? Comparison Corpus, Evaluation, and Detection". <https://arxiv.org/abs/2301.07597>

[^text-analysis-3]: David Robinson's analysis of the 2016 Trump twitter account: <http://varianceexplained.org/r/trump-tweets/>

### Getting the data

HC3 is **not** included in the **behdslabs** package. It is hosted by its authors on the Hugging Face Hub, it is fairly large, and part of the lesson here is learning to obtain and document a dataset you did not get from a package. The code below downloads four of the corpus's topic domains -- computer-science Wikipedia entries, open-domain quiz questions, finance, and medicine -- and skips the much larger `reddit_eli5` domain, whose text is subject to Reddit's terms of use. You only need to run the download once; the `if (!file.exists(...))` guard means re-running the chapter will reuse the local copies.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales)
library(tidytext)
library(textdata)
library(jsonlite)
```

```{r, echo=FALSE}
set.seed(2002)
```

```{r download-hc3}
#| cache: true
#| message: false
#| warning: false
domains <- c("wiki_csai", "open_qa", "finance", "medicine")
base_url <- "https://huggingface.co/datasets/Hello-SimpleAI/HC3/resolve/main/"

hc3_raw <- map_df(domains, function(domain) {
  file <- paste0(domain, ".jsonl")
  if (!file.exists(file)) download.file(paste0(base_url, file), file, quiet = TRUE)
  stream_in(file(file), verbose = FALSE) |> mutate(source = domain)
})
```

Each row is one question, and the human and ChatGPT answers are stored in *list columns* because a question can have more than one of each:

```{r}
glimpse(hc3_raw)
```

We want a table with one *answer* per row, labeled by who wrote it. We use `pivot_longer` to stack the two answer columns and then `unnest_longer` to expand the lists:

```{r}
answers <- hc3_raw |>
  transmute(question, source,
            human = human_answers, ai = chatgpt_answers) |>
  pivot_longer(c(human, ai), names_to = "author", values_to = "text") |>
  unnest_longer(text) |>
  filter(!is.na(text) & str_squish(text) != "") |>
  mutate(answer_id = row_number())
```

This leaves us with a good-sized corpus:

```{r}
answers |> count(author)
```

### A first look

Before touching individual words, one blunt summary already separates the two groups: **length**. Here is the distribution of answer lengths, in characters, for each author:

```{r answer-length-by-author}
answers |>
  mutate(n_char = str_length(text)) |>
  ggplot(aes(n_char, after_stat(density), fill = author)) +
  geom_histogram(binwidth = 100, boundary = 0, alpha = 0.5, position = "identity") +
  coord_cartesian(xlim = c(0, 3000)) +
  labs(x = "Answer length (characters)", y = "Density", fill = "")
```

```{r}
answers |>
  mutate(n_char = str_length(text)) |>
  group_by(author) |>
  summarize(median = median(n_char), sd = sd(n_char))
```

The ChatGPT answers are both **longer** and much more **uniform** in length: they cluster around a thousand characters, while the human answers range from one-line replies to long essays. A model that always writes a tidy three-paragraph answer looks different from a crowd of people, some terse and some verbose. This is our first hint that authorship leaves a measurable trace; next we look for it in the words themselves.

## Text as data

The **tidytext** package helps us convert free-form text into a tidy table. Having the data in this format greatly facilitates data visualization and the use of statistical techniques.

The main function needed to achieve this is `unnest_tokens`. A *token* refers to a unit that we are considering to be a data point. The most common *token* will be words, but they can also be single characters, n-grams, sentences, lines, or a pattern defined by a regex. The function takes a vector of strings and extracts the tokens so that each one gets a row in the new table. Here is a simple example:

```{r}
poem <- c("Roses are red,", "Violets are blue,",
          "Sugar is sweet,", "And so are you.")
example <- tibble(line = c(1, 2, 3, 4),
                  text = poem)
example
example |> unnest_tokens(word, text)
```

Now let's look at one of the answers. We pick a ChatGPT answer that contains a line-break marker directly against a word, because it will let us illustrate a cleaning problem:

```{r}
i <- which(answers$author == "ai" & str_detect(answers$text, "\\.\\\\n[A-Z]"))[1]
answers$text[i] |> str_wrap(width = 65) |> cat()
```

```{r}
answers[i, ] |>
  unnest_tokens(word, text) |>
  pull(word)
```

Note the odd tokens beginning with `n`, such as `nthe` or `nin`. The models format their answers into paragraphs and lists using the line-break marker `\n`; when `unnest_tokens` strips the backslash it leaves the `n` stuck to the next word. (A related issue: the answers also contain the HTML entities `&amp;`, `&gt;` and `&lt;`.) This is the analogue of stray links and picture URLs in social-media text, and we clean it the same way -- with a regular expression that removes these markers before tokenizing:

```{r}
clean_markup <- "\\\\[nrt]|&amp;|&gt;|&lt;"
answers[i, ] |>
  mutate(text = str_remove_all(text, clean_markup)) |>
  unnest_tokens(word, text) |>
  pull(word)
```

We are now ready to extract the words from all the answers.

```{r tokenize-answers}
#| cache: true
#| message: false
#| warning: false
answer_words <- answers |>
  mutate(text = str_remove_all(text, clean_markup)) |>
  unnest_tokens(word, text)
```

And we can now answer questions such as "what are the most commonly used words?":

```{r}
answer_words |>
  count(word) |>
  arrange(desc(n))
```

It is not surprising that these are the top words, which are not informative. The **tidytext** package has a database of these commonly used words, referred to as *stop words*:

```{r}
head(stop_words)
```

We remove the stop words, and we also drop tokens that are just numbers (using the regex `^\d+$`) and strip a leading `'` left over from quotations:

```{r}
answer_words <- answer_words |>
  filter(!word %in% stop_words$word & !str_detect(word, "^\\d+$")) |>
  mutate(word = str_replace(word, "^'", ""))
```

This leaves a much more informative set of top words:

```{r}
answer_words |>
  count(word) |>
  slice_max(n, n = 10) |>
  arrange(desc(n))
```

## Which words distinguish the authors?

For each word we want to know whether it is more likely to come from a human answer or a ChatGPT answer. We compute, for each word, its frequency among human words and among AI words, then take the ratio of these proportions (AI proportion divided by human proportion). Because some words are infrequent, we apply the continuity correction described in @sec-log-transform:

```{r}
human_vs_ai <- answer_words |>
  count(word, author) |>
  pivot_wider(names_from = "author", values_from = "n", values_fill = 0) |>
  mutate(p_ai = (ai + 0.5) / (sum(ai) + 0.5),
         p_human = (human + 0.5) / (sum(human) + 0.5),
         ratio = p_ai / p_human)
```

For words appearing at least 80 times in total, here are the ones most over-represented in the ChatGPT answers:

```{r}
human_vs_ai |> filter(ai + human >= 80) |> arrange(desc(ratio))
```

and the ones most over-represented in the human answers:

```{r}
human_vs_ai |> filter(ai + human >= 80) |> arrange(ratio)
```

A pattern emerges. The AI answers lean on a cautious, service-desk vocabulary -- *personalized*, *carefully*, *noting*, *alleviate*, *discomfort*, *experiencing*, *healthcare*. The human answers carry informal fillers (*pretty*, *basically*, *stuff*), first-person reasoning (*assuming*, *doubt*, *instance*), and artifacts of where the text came from: `http` links, the word *answered*, and `displaystyle` from LaTeX math in the Wikipedia-based answers. As in the original social-media case study, we are less interested in individual words than in the overall *tone*. For that we turn to sentiment analysis.

## Sentiment analysis

In sentiment analysis we assign a word to one or more "sentiments". This approach will miss context-dependent meaning, such as sarcasm, but applied to large numbers of words the summaries can be informative.

The **tidytext** package gives access to several sentiment lexicons through the **textdata** package. The `bing` lexicon divides words into `positive` and `negative`:

```{r, eval=FALSE}
get_sentiments("bing")
```

The `AFINN` lexicon assigns a score between -5 (most negative) and 5 (most positive). It has to be downloaded the first time you call the function:

```{r, eval=FALSE}
get_sentiments("afinn")
```

The `nrc` lexicon provides several finer-grained sentiments and also has to be downloaded the first time:

```{r}
get_sentiments("nrc") |> count(sentiment)
```

For our analysis we use the `nrc` lexicon:

```{r}
nrc <- get_sentiments("nrc") |> select(word, sentiment)
```

We combine words and sentiments with a join. Because many words carry more than one sentiment, we pass `relationship = "many-to-many"`:

```{r}
answer_words |>
  inner_join(nrc, by = "word", relationship = "many-to-many") |>
  select(author, word, sentiment) |>
  sample_n(5)
```

Now we compare the two authors by counting how often each sentiment appears in each group. For each sentiment we compute its proportion among all sentiment-tagged words for that author, and take the ratio (AI proportion divided by human proportion):

```{r}
sentiment_counts <- answer_words |>
  left_join(nrc, by = "word", relationship = "many-to-many") |>
  count(author, sentiment) |>
  pivot_wider(names_from = "author", values_from = "n") |>
  mutate(sentiment = replace_na(sentiment, "none"),
         p_ai = ai / sum(ai), p_human = human / sum(human),
         ratio = p_ai / p_human) |>
  arrange(desc(ratio))
sentiment_counts
```

The differences are smaller than in the word table -- both authors are, after all, answering the same questions -- but the direction is consistent. Human answers are relatively richer in **surprise**, **joy**, **anger**, and **anticipation**: the vocabulary of someone reacting to a question. The ChatGPT answers skew toward **fear** and **sadness** words (which include a lot of clinical and cautionary terms like *symptoms*, *risk*, *pain*) and, overall, read as more uniformly measured. An LLM tuned to be a calm, comprehensive assistant flattens the emotional range that a room full of people would show.

To see which words drive these differences, we go back to the `human_vs_ai` object. For each sentiment we show the words with the largest ratios in either direction, excluding words appearing fewer than 10 times total:

```{r sentiment-words-by-author, out.width="100%", echo=FALSE}
human_vs_ai |>
  filter(ai + human >= 10) |>
  inner_join(nrc, by = "word") |>
  mutate(sentiment = factor(sentiment, levels = sentiment_counts$sentiment)) |>
  group_by(sentiment) |>
  slice_max(abs(log(ratio)), n = 10) |>
  ungroup() |>
  mutate(word = reorder(word, -ratio)) |>
  ggplot(aes(word, ratio, fill = ratio < 1)) +
  geom_col(show.legend = FALSE) +
  scale_y_log10() +
  facet_wrap(~sentiment, scales = "free_x", nrow = 2) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

This is just a small example of what can be done with **tidytext**. To learn more we recommend the *Text Mining with R* book[^text-analysis-4] by Julia Silge and David Robinson.

[^text-analysis-4]: <https://www.tidytextmining.com/>

## Exercises

Project Gutenberg is a digital archive of public domain books. The R package **gutenbergr** facilitates the importation of these texts into R.

You can install and load it by typing:

```{r, eval=FALSE}
install.packages("gutenbergr")
library(gutenbergr)
```

You can see the books that are available like this:

```{r, eval=FALSE}
gutenberg_metadata
```

1\. Use `str_detect` to find the ID of the novel *Pride and Prejudice*.

2\. We notice that there are several versions. The `gutenberg_works()` function filters this table to remove replicates and include only English language works. Read the help file and use this function to find the ID for *Pride and Prejudice*.

3\. Use the `gutenberg_download` function to download the text for *Pride and Prejudice*. Save it to an object called `book`.

4\. Use the **tidytext** package to create a tidy table with all the words in the text. Save the table in an object called `words`.

5\. We will later make a plot of sentiment versus location in the book. For this, it will be useful to add a column with the word number to the table.

6\. Remove the stop words and numbers from the `words` object. Hint: use the `anti_join`.

7\. Now use the `AFINN` lexicon to assign a sentiment value to each word.

8\. Make a plot of sentiment score versus location in the book and add a smoother.

9\. Assume there are 300 words per page. Convert the locations to pages and then compute the average sentiment in each page. Plot that average score by page. Add a smoother that appears to go through the data.
